---
title: "Bison"
author: "Socorro Dominguez"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    fig_caption: yes
    keep_md: yes
    self_contained: yes
    theme: readable
    toc: yes
    toc_float: yes
  pdf_document:
    pandoc_args: "-V geometry:vmargin=1in -V geometry:hmargin=1in"
---

```{r setup, echo=FALSE}
options(warn = -1)
pacman::p_load(neotoma2, dplyr, tidyverse, ggplot2, sf, geojsonsf, leaflet, DT, maptools, maps)
```

## 1. Exercise Using Vertebrate Data

Let's retrieve Bison data and see how they were spread out during the Holocene and Pleistocene.

### 1.a Getting "Bison" Data from Neotoma

1. To get the sites that contain `bison` data, let's use the `get_sites()` function and use the parameter `taxa`. Remember that we can use the wildcard `%` to retrieve "similar" names. To retrieve **all** the data, set the `all_data` parameter to TRUE.

```{r finding bison}
# Might take a few minutes...
bison <- get_sites(taxa="%bison%", all_data=TRUE)
head(bison)
```

2. Now that we have downloaded the sites metadata, let's use that to see where are `bison` located:

```{r plot bison sites}
plotLeaflet(bison)
```

3. Now, let's get all the rows of sample data in Neotoma. Since there are 1258 sites, this operation might take a while, so an .RDS file `bisonDataDL.RDS` has been provided. Let's load that:

```{r obtain records}
bison_dl <- readRDS('data/BisonDataDL.rds')
head(bison_dl)
```

3.a. What if we didn't have the `.RDS` file? The code we would have had to use is:
```{r download the data from scratch}
#bison_dl <- get_downloads(bison)
```

4. Let's take a look at the samples in the `bison_dl`.
You might notice that there are variable names that do not belong to `bison`, that is because the dataset type is `vertebrate-fauna` and the `get_downloads()` function downloads data at the `datasetid` level. We will filter information and keep only the `bison` data by joining with a translation table that contains the `variablename`s we would be interested in.

```{r getSamples}
allSamp <- samples(bison_dl)
```
```{r view Samples}
DT::datatable(head(allSamp))
```

5. Let's load in the "translation" table for bison (`bisontable.csv` in the data folder) and take a look at it:

```{r translationDisplay}
translation <- readr::read_csv("data/bisontable.csv")
DT::datatable(translation, rownames = FALSE, 
                options = list(scrollX = "100%", dom = 't'))

```

6. Now, we want to join based on the `variablename` but we just want to keep the `harmonizedname`.

```{r join tables}
allSampDF <- allSamp %>%
  inner_join(translation, by = c("variablename" = "variablename")) %>%  # Join by `variablename`
  dplyr::select(!c("variablename")) # Remove `variablename`, we only care about the `harmonizedname` for now.
DT::datatable(head(allSampDF))
```
7. Let's look how many rows we have left:
```{r}
nrow(allSampDF)
```
7. In order to visualize how many `bison` were at a a particular time period and location, let's group them by `siteid`, `sitename`, `harmonizedname`, `sampleid`, `units`, `ageyounger`, `ageolder`, `agetype`, `depth`, `datasetid`, `long`, `lat`; I will also sum the `value` or number of observations.

It is probably not the best way to group them by, but for purposes of this coding example, that is how I will do it.
``` {r}
allSampDF <- allSampDF %>% 
  group_by(siteid, sitename, harmonizedname,
           sampleid, units, ageyounger, ageolder, 
           agetype, depth, datasetid,
           long, lat) %>%
  summarise(value = sum(value), .groups='keep')
DT::datatable(head(allSampDF))
```
```{r nrows II}
nrow(allSampDF)
```

8. Let's add a new column where we define if some of the given observations belong to the `Pleistocene` or `Holocene`. We will use `ageyounger` for this purpose. Let's remove the observations that have no era defined.

```{r pleistocene and holocene}
allSampDF <- allSampDF %>% 
  mutate(era = factor(ifelse(ageyounger >= 11700, "Pleistocene", "Holocene"))) %>% # Assign a new column with the values Pleistocene or Holocene based on ageyounger
  mutate(era  = fct_relevel(era,"Pleistocene", "Holocene")) %>%
  filter(!is.na(era)) # Remove empty eras

DT::datatable(head(allSampDF))
```


```{r nrows III}
nrow(allSampDF)
```
9. Let's visualize a quick bar plot that counts how many observations belong to each era.

```{r plotting data}
taxaplots <- allSampDF %>%
  group_by(era, harmonizedname) %>%
  summarise(sites = length(unique(siteid)), samples = length(unique(sampleid)), .groups='keep') %>% 
  arrange(era)

DT::datatable(taxaplots)
```

```{r bison sites in each era}
ggplot(data = taxaplots, aes(x = era, y=samples)) +
  geom_bar(stat = 'identity') +
  ggtitle("Bison Count per Era") +
  xlab("Age") +
  ylab("Number of Bison") +
  theme_bw()
```
Let's plot the observations of each era on the map; we will use the `map_data` function from the `maps` library:

```{r}
world <- map_data("world")
```

10. Let's show the observations using a facet plot:

```{r facetPlot}
ggplot(allSampDF, aes(x = long, y = lat)) + 
              geom_map(data = world, 
                       map = world,
                       aes(map_id = region),
                       color = "black",
                       fill = "white",
                       show.legend = FALSE,
                       inherit.aes = FALSE) +
              geom_point(aes(colour=era), show.legend = FALSE) + 
              facet_grid(era ~ .) +
              theme_bw() +
              ggtitle("Bisons in each Era") +
              xlab("Longitude") + 
              ylab("Latitude") 
```