---
title: "SVP 2016"
author: "Jessica Blois, Edward Davis, Simon Goring, Jack Williams, Eric C. Grimm"
csl: ecology.csl
output:
  html_document:
    highlight: pygment
    includes:
      after_body: styles/footer.html
      before_body: styles/header.html
    keep_md: yes
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    reference_docx: styles/word-styles-reference-01.docx
    toc: yes
    toc_depth: '3'
bibliography: workshop.bib
---

# Introduction

This workshop will provide 1) guidance on best practices in the archiving and analysis of paleovertebrate data and 2) training in the use of the Neotoma Paleoecology Database (www.neotomadb.org) to archive, access, and analyze paleoecological data. Neotoma is a multiproxy paleodatabase that stores multiple kinds of paleoecological & paleoenvironmental data, including vertebrate faunal data. One of the strengths of Neotoma is the ability to compare faunal data with other proxy data such as fossil pollen, diatoms, ostracodes, insects, charcoal, and geochemical data.  In addition, the database is structured to relate absolute dates to taxon occurrences and to allow the creation and storage of age models built on absolute dates from stratigraphic sections. Neotoma is a public-access, community-supported database that is emerging as the standard repository for Pliocene and Quaternary paleoecological data.

More teaching materials can be found [here](http://www.neotomadb.org/education/category/higher_ed/)

# Finding Data

## Explorer

### Getting Started
  1. Go to [http://www.neotomadb.org/](http://www.neotomadb.org/) and click on the 'Explorer' picture, or navigate directly to the [Explorer App](http://apps.neotomadb.org/Explorer/)
  2. Pan (by dragging), or change the zoom so that your window is centered on North America, including all of the lower 48 states and the southern half of Canada.

###	Tips and tricks
  1. Show/Hide Search Results
    a. You often accumulate many search layers - it may be getting a bit confusing.  Find the icon that lets you show/hide/combine search layers and use it to hide or delete some of your searches.
  2. Rename searches
    a. Searches from the "Basic" search window are automatically named, but not "Advanced" searches.  Add names for your searches as you go at the bottom of the Search dialog

### Search for Data

####	Finding sites
  1. **Find a known site**
    a. Using the Search dialog window (Advanced tab, Metadata subtab, Site Name field), find the classic Guilday site "New Paris #4". 
      i. *Note:* Search doesn't like the # symbol, so search for "New Paris", which should return two sites: New Paris #2, and New Paris #4
    b. Once you've performed the search, click on the point that appears.  A window will pop up with some information about this record.
    
    > QUESTION 1:  What is the latitude and longitude of New Paris #4?  What is the Site ID?
  
  2. **Explore sites by geography**
    a. Using the Search dialog window (Advanced tab), first choose "dataset type = vertebrate fauna" at the top
    b. Then, in the Space subtab, click "Search by extent"
    c. Click the "Extent" dropdown menu and search by shape, select the rectangle, then draw a rectangle on the map in your chosen region of North America
    
    > QUESTION 2:  How many total sites are found in that region?

  3. **Find all sites produced by a researcher**
    a. Search Window, Advanced Tab, Metadata subtab, Person Name field
    b. Find all sites produced by Ernie Lundelius
    
    > QUESTION 3:  In which states has Ernie worked?

####  Find a Taxon 
  1. **Search for a single taxon.**
    a. Search Window, Basic Tab, use Taxon field
    b. Find all sites with *Antilocapra* records
  
    > QUESTION 4:  How many sites have *Antilocapra* records?
  
  2. **Search for multiple taxa**
    a. The basic *Antilocapra* search indicates that we might need to build up the taxonomy.  Let's re-search for *Antilocapra*, but make sure we've included all relevant records.
    b. Search Window, Advanced tab, Taxa subtab
      i. *Note:* To do this, we will search for a "Taxa group".To the right of the "Taxon name" field, click on the gear symbol. Click on "Mammals" for the Taxa group, then search for *Antilocapra*. Then, click on all taxa you want to include here.  For example: Antilocapra americana, Antilocapra sp., Antilocapra, cf. Antilocapra americana, cf Antilocapra sp., Antilocapra cf A. americana, Antilocapridae ?Antilocapra sp., and ?Antilocapra sp. 
    
    > QUESTION 5:  What state has the easternmost location of *Antilocapra* in this search?  (For comparison, the eastern range limit of *Antilocapra* today is in western Iowa).
    
    c.  Add the modern species range onto the search: Click on the red polygon at the top, search for and select *Antilocapra americana*
  
  
####	Find all vertebrate records in North America 
  1. **Search window, Advanced Tab, Dataset type = "Vertebrate fauna"**
  
  > QUESTION 6:  The generation of fossil records is labor intensive and hence expensive - *e.g.* the costs of fieldwork, the money spent on radiocarbon dates, the time required for a trained analysis to identify specimens, etc. A rough time/cost estimate for a single vertebrate fossil record is on the order of two years and $30,000.  Given this, give an order-of-magnitude estimate of the number of person-years and dollars it took to generate these fossil vertebrate records now stored in Neotoma. (Order-of-magnitude = 10 person-years?  100 person-years? 1000? etc.)
  
    
####	Multi-Taxon Search
  1. **Find all sites with at least 20% *Picea* pollen between 15,000 and 12,000 years ago.**
    a. For *Picea*, first click on 'Dataset Type" = "pollen"
    b. Then use the 'Advanced Taxon Selection' which you can use by clicking on the gears icon to the right of the 'Taxon' field in the 'Search' window. 
      i. Enter 'Taxa group' = Vascular Plant, search for Picea,click the box next to Taxon to check all taxa, then uncheck taxa  like "Picea/Pinus" or "Picea/Abies".
    c. Then click the 'Abundance/density' box and select >20%.  
    d. Fill in the appropriate age range, and choose records that 'intersects result age range'. 
    e. Finally, click 'Search'
  2. **Find all sites with *Mammut* (mastodon) between 15,000 and 12,000 years ago.**
    a. For *Mammut*, use the 'Advanced Taxon Selection' which you can use by clicking on the gears icon to the right of the 'Taxon' field in the 'Search' window.
      i. In the 'Advanced Taxon Selection' window, choose 'Mammals' for Taxa Group and then enter *Mammut* into the 'Search for' window.Then click 'Go'
      ii. Note that the search returned taxon names for both *Mammut* (mastodon) and *Mammuthus* (mammoth).  Click all boxes for all variants of *Mammut* but do not click the *Mammuthus* boxes
      iii. Enter a search name (e.g. '*Mammut* - all') and click Save
      iv. In the general search window, click 'Search'
      
      > QUESTION 7:  Take a screenshot of this map and include it in your homework.  Does mastodon tend to live in places with spruce, or without spruce?  Suggest two hypotheses that might explain the observed association (or lack of one).
      
####	Multi-Time Search
  1.	Hide your previous searches.
  2.	Find all sites with *Mammuthus* between 21,000 and 18,000 years ago.
  3.	Find all sites with *Mammuthus* between 18,000 and 15,000 years ago.
  4.	Find all sites with *Mammuthus*  between 15,000 and 13,000 years ago.
  5.	Find all sites with *Mammuthus*  between 13,000 and 11,000 years ago.
  6.  Find all sites with *Mammuthus*  between 11,000 and 8,000 years ago.
  
  > QUESTION 8:  Take a screenshot of this map and include it in your homework.  Describe the history of *Mammuthus* distributions in North America over the last 21,000 years.   

### View Data

  1.	Using the last *Mammuthus* search (between 11,000 - 8,000 years ago), click on the single site that appears (Fetterman Mammoth Locality).  
  2.	In the popup window with metadata for Fetterman, note that the bottom includes a list of datasets available at the site.  (V = vertebrate; clock = geochronological data)
  3.	Hold the mouse over the vertebrate dataset.  Note that an eyeball and '+' appear to the right.
  4.	Click on the eyeball to view the Fetterman Mammoth Locality vertebrate dataset.  This opens up a new window with more detail about your dataset.
  a. Explore the specimens found at the site. There is no associated geochronological dataset, but the chronology tab indicates an age for the mammoth. What other information would you want about this date to determine authenticity?
    + Samples:  A data table.  Each row is a different variable, the first few columns show metadata associated with each specimen, and each subsequent column is a stratigraphic depth.  In this case, this is an assemblage so there is only one depth column.
    + Site: General information about the site
    
    > QUESTION 9:  What publications are listed for Fetterman?
    
  5.  Now search for the site "Samwell Cave".  
    + Notice that this site has multiple localities within it, and each locality has both a vertebrate dataset and a geochronological dataset.
    a. Explore the Samwell Cave Popcorn Dome datasets further.
    
    > QUESTION 10: What is the difference between information in the "Chronology" tab of the Vertebrate dataset versus the information within the Geochronology dataset?


### Download Data
  1.	In the popup window with metadata for Samwell Cave Popcorn Dome, hold the mouse over the vertebrate dataset.  Note that an eyeball and '+' appear to the right.  Click on the '+'.  This adds the dataset to a 'Datasets' tray.
  2.	Find the icon for the 'Saved Datasets Tray' and click on it.
  3.	Hover over the dataset then click on the Save icon to the right.  The dataset will be saved as a text file in CSV (comma separated value) format.
  4.	Open the downloaded CSV file in Excel or a text editor (e.g. Notepad, Wordpad) to look at it.
  
  > QUESTION 11:  When was the last time *Aplodontia rufa* was seen at the site?

## The `neotoma` Package

```{r}
# Uncomment this line if you haven't already installed any of these packages:
# install.packages(c("neotoma", "analogue"))

#Add the neotoma package to your programming environment (we'll add analogue later)
library(neotoma)

```

`neotoma` has three fundamental commands: `get_site`, `get_dataset`, `get_download`. The first two return metadata for sites and datasets; the latter returns data. See Goring et al. [@neotoma_goring] for a full description of the package and example code.  This exercise is partially based on those examples.

### Finding sites

We'll start with `get_site`.  `get_site` returns a `data.frame` with metadata about sites. You can use this to find the spatial coverage of data in a region (using `get_site` with a bounding box), or to get explicit site information easily from more complex data objects.  Use the command `?get_site` to see all the options available.

You can easily search by site name, for example, finding "Marion Lake".  By default the search string is explicit, but because older sites, entered as part of COHMAP, often had appended textual information (for example `(CA:British Columbia)`), it's often good practice to first search using a wildcard character:

```{r}

marion_site <- get_site(sitename = 'Marion%')

print(marion_site)
```

While `marion_site` is a `data.frame` it also has class `site`, that's why the print output looks a little different than a standard `data.frame`.  That also allows you to use some of the other `neotoma` functions more easily.  

```{r}

#Search by lat/lon bounding box.  This one roughly corresponds to Florida.
FL_sites <- get_site(loc = c(-88, -79, 25, 30)) 
```

You can also search by geopolitical name or geopolitical IDs (`gpid`) stored in Neotoma. For a list of names and gpids, go to [http://api.neotomadb.org/apdx/geopol.htm](), or use the `get_table(table.name = "GeoPoliticalUnits")` command.  This command works either with an explicit numeric ID, or with a text string:

```{r}
#get all sites in New Mexico (gpid=7956)
NM_sites <- get_site(gpid = 7956)

#get all sites in Wisconsin
WI_sites <- get_site(gpid = "Wisconsin")
```

`data.frame`s store vectors of equal length.  The nice thing about `data.frame`s is that each vector can be of a different type (character, numeric values, *etc*.). In RStudio, you can use the Environment panel in upper right to explore variables. 

We pointed out before that the object returned from `get_site` is both a `data.frame` and a `site` object.  Because it has a special `print` method some of the information from the full object is obscured when printed.  You can see all the data in the `data.frame` using `str` (short for *structure*):

```{r}
str(marion_site)
```

Let's look at the `description` field:

```{r}
marion_site$description
```


### Getting Datasets

The structure of the Neotoma data model, as expressed through the API is roughly: "`counts` within `download`s, `download`s within `dataset`s, `dataset`s within `site`s".  So a `dataset` contains more information than a site, about a particular dataset from that site.  A site may have a single associated dataset, or multiple.  For example:

```{r}

get_dataset(marion_site[1,])

```

```r
#get_dataset returns a list of datasets containing the metadata for each dataset
#We can pass output from get_site to get_dataset
marion.meta.dataset  <- get_dataset(marion.meta.site)
#Let's look at the metadata returned for Marion Lake and Marion Landfill.  Both
#have a geochronology dataset, while one has a pollen dataset and the other a 
#vertebrate fauna dataset
marion.meta.dataset
```

### Get_Download

`get_download` returns a list which stores a list of download objects - one for each retrieved dataset.  Each download object contains a suite of data for the samples in that dataset.  Get all datasets for both Marion Site and Marion Landfill. `get_download` will accept an object of class dataset:

```{r}

marion_all <- get_download(marion_site)

print(marion_all)
```

There are a number of messages that appear.  These should be suppressed with the flag `verbose = FALSE` in the function call.  One thing you'll note is that not all of the datasets can be downloaded directly to a `download` objct.  This is because `geochronologic` datasets have a different data structure than other data, requiring different fields, and as such, they can be obtained using the `get_geochron` function:

```{r}

marion_geochron <- get_geochron(marion_site)

print(marion_geochron)
```

The result is effectively the inverse of the first.

```{r}
#Get all datasets for just Marion Lake (BC):
marion_bc <- get_download(marion_site[1,])
```

Within the download object, `sample.meta` stores the core depth and age information for that dataset. We just want to look at the first few lines, so are  using the head function.

```{r}
head(marion_bc[[1]]$sample.meta)

#taxon.list stores a list of taxa found  in the  dataset
head(marion_bc[[1]]$taxon.list)

#counts stores the the counts, presence/absence data, or percentage data for each taxon for each sample
head(marion_bc[[1]]$counts)

#lab.data stores any associated  laboratory measurements in the dataset
#For Marion Lake, this returns the Microsphere suspension used as a spike to calculate
#concentrations
head(marion_bc[[1]]$lab.data)
```

### Helper functions

#### `compile_taxa`

The level of taxonomic resolution can vary among analysts.  Often for multi-site analyses it is helpful to aggregate to a common taxonomic resolution. The `compile_taxa` function in `neotoma` will do this.  To help support rapid prototyping, `neotoma` includes a few pre-built taxonomic lists, **however**, the function also supports the use of a custom-built `data.frame` for aligning taxonomies.  Because new taxa are added to Neotoma regularly (based on analyst identification), it is worthwhile to check the assignments performed by the `compile_taxa` function, and to build your own explicit compilation table.

```{r}
marion_bc <- compile_taxa(marion_bc, list.name = "P25")
```

You'll notice that warning messages return  a number of taxa that cannot be onverted using the existing data table.  Are these taxa important?  They may be important for you.  Check to see which taxa have been converted by looking at the new taxon table:

```{r}
marion_bc[[1]]$taxon.list[,c("compressed", "taxon.name")]
```

And note that if you look at the names of the objects in the new `download` (using `names(marion_bc[[1]]))`, there is now a `full.counts` object.  This allows you to continue using the original counts, while also retaining the new compiled counts.

#### Plotting

There are several options for plotting stratigraphic data in R.  The `rioja` package [@rioja_package] and `analogue` [@analogue_package] each have methods, and other possibilities exist.  Here we will show simple plotting using the `analogue` package. To make it clear which functions come from the `analogue` package I will use `analogue::` before the function names.  This is just an explicit way to state the function source.  If you choose not to do this you will not encounter any problems unless multiple packages have similarly name functions.

```{r, message=FALSE, warning=FALSE}

library("analogue")

# Convert the Marion Lake pollen data to percentages
marion_bc_pct <- analogue::tran(x = marion_bc[[1]]$counts, method = 'percent')

# Drop rare taxa:
marion_bc_pct <- marion_bc_pct[, colMeans(marion_bc_pct, na.rm = TRUE) > 2]

analogue::Stratiplot(x = marion_bc_pct[ , order(colMeans(marion_bc_pct, na.rm = TRUE), 
                                                decreasing = TRUE)], 
                     y = marion_bc[[1]]$sample.meta$age,
                     ylab = marion_bc[[1]]$sample.meta$age.type[1],
                     xlab = " Pollen Percentage")

```

# Age Models

We can use the Neotoma package, along with other tools to rebuild age models. Many of the age modeling tools require additional files to be installed.  One such program is [Bacon](http://chrono.qub.ac.uk/blaauw/bacon.html) [@].  Here we will rebuild a Baysian age model for a site in Wisconsin.

We've looked in detail at Marion Lake, let's choose another site from our pool of Wisconsin sites earlier.  First we need to get Bacon set up:

*Jessica Note: section deleted from this document.  See template to recover*

# Multi-Site Analysis

# Conclusions

# References